Interactive Visualization of Hierarchically Structured Data

### 2017-03-23 treelapse.Rmd
Write description for structSSI

[@dethlefsen2008pervasive] investigates the effect of antibiotics on
bacterial community composition from an ecological perspective. The
study tracks the microbiome of three patients across ten months, with
two five-day antibiotic time courses separated by 6 months. Discerning
the variation in resilience across bacteria is important, considering
the the role of bacteria in health and not just disease.

@@

The study [@digiulio2015temporal] tracked the abundance of bacteria in
the vaginal microbiome during pregnancy in an effort to study
relationships between bacterial community composition and preterm
birth. Ideally, it would be possible to develop clear bacterial
signatures associated with preterm births.

@@

>>>>

@dethlefsen2008pervasive investigated the effect of antibiotics on
bacterial community composition from an ecological perspective. The
study tracks the microbiome of three patients across ten months, with
two five-day antibiotic time courses separated by 6 months. Discerning
the variation in resilience across bacteria is important, considering
the the role of bacteria in health and not just disease.

@@

@digiulio2015temporal tracked the abundance of bacteria in
the vaginal microbiome during pregnancy in an effort to study
relationships between bacterial community composition and preterm
birth. Ideally, it would be possible to develop clear bacterial
signatures associated with preterm births.

@@

### Inspecting Confirmatory Analysis {#structssi}

In addition to facilitating exploratory study, treelapse has potential value as
a device for inspecting confirmatory analysis. We provide an illustration
extending an example from [@callahan2016bioconductor], which formally tested
bacterial species for association with age in a sample of mice. The testing
approach advocated there is particularly well-suited to visualization with
treelapse, since it sought to detect associations at multiple levels of
phylogenetic resolution, using statistical tools developed in
[@yekutieli2008hierarchical; @sankaran2014structssi].

The data of interest in [@callahan2016bioconductor] are bacterial counts
collected across old and young mice. After variance-stabilizing these counts
using DESeq2 [@love2014moderated], a $t$-test was applied to each node in a
phylogenetic tree, comparing abundances between old and young mice. To account
for multiple testing, we employ the structSSI algorithm
[@yekutieli2008hierarchical; @sankaran2014structssi] along with methods
available in the `multtest` package [@pollard2005multiple].

To interpret the results, we apply timebox trees. Our goals are to (1) identify
subtrees with consistently elevated differential abundance across age groups
and (2) compare alternative multiple testing adjustment procedures. Our approach
is to display the negative-log raw and adjusted $p$-values for each node, with
alternative methods compared via parallel coordinates. One view of the resulting
display is captured in Figure \@ref(fig:structssi). First, we see that
significant nodes tend to be significant across all methods -- the ordering
between different series appears stable. Interestingly, the Sidak one-step and
structSSI procedures seem to have lower power than the others, including
conservative FWER-controlling methods, like the Bonferroni procedure. Further,
in this application, FDR-controlling techniques do not seem to offer notably
different adjusted $p$-values, relative to those controlling FWER. This suggests
that, for this problem, bacteria are either strongly associated with age, or not
associated at all, so that there is little gain from using more sensitive
procedures.

```{r structssi, fig.cap = "Viewing a tree of $p$-values across different methods highlights two subtrees with strong associations with mouse age, across several testing procedures."}
include_graphics("figure/structssi.png")
```

Further, selecting series with strong association between abundance and age, two
major subtrees are brought to the forefront. Separately querying the the
taxonomic identities of these bacteria reveals that they are two subgroups of
Clostridia, which is consistent with the analysis of
[@callahan2016bioconductor]. More than this specific analysis outcome, this view
demonstrates that interactive visual inspection of results from confirmatory
analysis provides deeper insight than the standard practice of printing tables
of (adjusted or unadjusted) $p$-values: the relationship between significant
nodes is only clear upon visualization on the tree.
